---
#title: "Bayesian mixed models, shrinkage, sparse and interaction kernel regression"
#author: "Luc Janss, Marni Tausen, Maria Izabel Cavassim Alves, Zhe Zhang"
#date: "`r Sys.Date()`"
output:
  md_document:
    variant: markdown_github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bayesian mixed models, shrinkage, sparse and interaction kernel regression
[![Build Status](https://travis-ci.org/ljanss/Rbayz.svg?branch=master)](https://travis-ci.org/ljanss/Rbayz)[![Coverage Status coveralls](https://coveralls.io/repos/github/ljanss/Rbayz/badge.svg?branch=master)](https://coveralls.io/github/ljanss/Rbayz?branch=master)

Jump to:

- [R/bayz overview and features](#R/bayz overview and features)
- [Downloading and installing R/bayz](#Downloading and installing R/bayz)
- [A very short tour](#A very short tour of R/bayz)
- [Further help](#Further help)

# R/bayz overview and features

R/bayz -- package Rbayz -- with main function bayz():

- covers mixed models in a Bayesian framework with uncertainty estimates on coefficients and variance parameters,
  MCMC convergence diagnostics, HPD intervals and trace plots.
- allows to add kernels (relationship / similarity matrices) on random effects, and multiple kernels
  on interactions, fitted as reduced rank tensors.
- can learn sparse multi-dimensional kernels as tensor decompositions, smooth kernels, sparsify kernels,
  and borrow information across kernels.
- directly models large sets of multiple features with different shrinkage regression options, and can obtain
  significances for individual features as p-values from random effects.
- offers standard R methods for summary, extraction of estimates, plot, predict and contrasts.
- is computationally efficient and can be used on large datasets, with parallelization options for MCMC sampling.

R/bayz has been developed in agriculture and ecology for various modeling and prediction needs using genomics, multi-omics,
phenomics, enviromics, time-series images and spectral data, multi-time multi-tissue transcriptomics data,
and more. This includes basic applications such as genomic and genomic-by-enviromic prediction and multi-trait modeling using
a Bayesian factor-analytic approach,
is extendable to efficienctly learn covariance structures in higher order interactions such as trait x time x enviroment,
and may be useful in any domain where mixed models, (multiple) kernels and/or large sets of predictors are relevant.

For further help check links in the text above, and see the overview of manual pages below. 

# Downloading and installing R/bayz

R/bayz us not yet released as a cran R package. It can be downloaded and installed with the options
shown below.

### 1. Precompiled binary versions (Windows, Mac)

There are precompiled binary packages for Windows and MacOS, which can be installed using install.packages() or with
devtools-tools. The MacOS precompiled version is built on Mac silicon architecture, the Windows version on Windows 11.
If these precompiled versions do not work
on your Windows or MacOS system, use option 2 below to install from github source.

Using install.packages() requires to use repos=NULL to circumvent use of cran, but which requires then to manually
install dependencies:
```{r, eval=FALSE}
install.packages("Rcpp")
install.packages("nlme")
install.packages("coda")
install.packages("https://ljanss.github.io/Rbayz/Rbayz_0.9.0.zip", repos=NULL, type="win.binary") # Windows
install.packages("https://ljanss.github.io/Rbayz/Rbayz_0.9.0.tgz", repos=NULL, type="mac.binary") # MacOS
```

Alternatively, if you have devtools installed, you can use devtools::install_url(), which automatically does install dependencies as well: 
```{r, eval=FALSE}
devtools::install_url("https://ljanss.github.io/Rbayz/Rbayz_0.9.0.zip", type="win.binary") # Windows
devtools::install_url("https://ljanss.github.io/Rbayz/Rbayz_0.9.0.tgz", type="mac.binary") # MacOS
```

### 2. Install source from Github repository (all systems including linux)

This requires a development environment in R, which needs Rtools on Windows, or "command line tools" on MacOS,
and the devtools package. On linux the development tools may often be pre-installed. The below commands run
in the R terminal.

Installing devtools and package dependencies:
```{r, eval=FALSE}
install.packages("devtools")
install.packages("Rcpp")
install.packages("nlme")
install.packages("coda")
```

Download and install/compile Rbayz using:
```{r, eval=FALSE}
library(devtools)
devtools::install_github("ljanss/Rbayz")
```

# Further help

R/bayz has a single main function bayz() that accepts model formulas in an extended
R-formula syntax where all
explanatory (right-hand-side) variables are wrapped by a "function-like" to specify
how to fit it in the model. 
This may look like Yield ~ fx(Year) + rn(Variety)
to fit Yield with Year as a fixed factor and Variety as a random factor (fx() and rn()
imply the variable used should be (converted to) factor).
For random effects, variance structures can be added with a V= option inside the
rn(), for instance with a given relationship / similarity matrix rn(Variety,V=Gmat),
which implies the variance-covariance structure Gmat $\sigma_g^2$.

Interactions between fators are specified using the colon, such as fx(Year:Location)
for fixed effects of Year-Location interactions. R/bayz does
not support automatic expansion with main effects, and main effects, if desired, should
be explicitly added in the model. For interactions in random effects, the variance specification
is expanded to a product of terms (interpreted as Kronecker product), written as
rn(Variety:Environment, V=Gmat*Emat), implying the variance-covariance structure for
the interaction effects as Gmat $\otimes$ Emat $\sigma_{ge}^2$. Using interactions
and Kronecker products involving estimated unstructured covariance matrices allows to
fit multi-trait models by supplying data in a "melted" form and specifying interactions
with Trait. Interactions and the matching variance structures can be specified to any order.
R/bayz can additionally fit large sets of predictors using the rr() (ridge or random regression)
model function, regular fixed (nested) regressions using rg(), and random slope models using rs().

R/bayz is used as many other R model functions by capturing the output object, and running methods
on that output like summary(), plot(), various methods to extract part of the estimates, and a
contrast and predict method. The summary() gives MCMC convergence diagnostics and Highest Posterior
Density regions for selected "traced" parameters, while plot() produces trace plots and density plots.

Check the following manual pages for additional information on:

-   An extended tour of R/bayz with examples
-   Full syntax for the bayz() function call
-   Overview of all model-function terms and their options
-   Overview of all variance-model specifications
-   Specification of residual variance structures
-   Using the output with summary() and other methods

# R/bayz technical details

R/Bayz is implemented in C++ using Rcpp. 

Current version: 0.12 (March 2026).

